\begin{abstract}
% 中文: LLM 在不同代码原语上表现出异质性，但 accuracy 无法解释内部差异来源。我们区分
% "信息已存在"与"信息已可用"，构建 dual diagnostic framework 追踪代码推理的内部生命周期。
% 发现 brewing 过程和四种 resolution outcome，不同 code primitive 诱导不同 fingerprint，
% brewing scaffold 跨模型稳定而 outcome 随能力变化。

LLMs show striking heterogeneity on code reasoning: the same model may succeed on variable tracking yet fail on loops or conditionals. Task accuracy alone does not explain these differences, because similar outputs can arise from different internal failures. We therefore distinguish \textit{information availability} from \textit{information readiness}: whether the correct answer is present in a hidden state, and whether the model can stably decode it itself. To study this gap, we introduce a dual diagnostic framework that combines layer-wise linear probing with Context-Stripped Decoding, and apply it to six synthetic code task families across five model families ranging from 0.5B to 14B parameters. We find a consistent internal lifecycle. Models first pass through a \textbf{brewing} stage, where the answer is linearly recoverable before it becomes self-decodable, and then enter one of four causally validated \textbf{resolution outcomes}: Resolved, Overprocessed, Misresolved, or Unresolved. The framework reveals that most code primitives are predominantly Resolved, but Computing---which combines multi-hop variable tracking with arithmetic---stands out with a Resolved rate below 34\%, and is the only task where Misresolved and Unresolved outcomes appear in substantial proportions. Controlled difficulty sweeps further show that different primitives harbor distinct latent failure bottlenecks that surface under increased task complexity. Across architectures and scales, the brewing scaffold remains stable while resolution success changes, suggesting that mechanism and capability are partly separable.
\end{abstract}
